---
title: "Alcohol attributable fraction lookup table sources"
author: "Ian Bates"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        #css: mystyle.css
vignette: >
  %\VignetteIndexEntry{Alcohol attributable fraction lookup table sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r global, include = FALSE}
library("aafractions.ncc")
```

## Introduction

The tables presented here are taken from online publically available sources.  These sources are listed here.

## Alcohol attributable fraction lookup table sources

### Current

[Local Alcohol Profiles for England 2017 user guide (.pdf)](https://fingertips.phe.org.uk/documents/LAPE_2017_User_Guide_071117.pdf) ([local link](`r system.file("aa_documentation", "LAPE_2017_User_Guide_071117.pdf", package = "aafractions.ncc")`))

* Appendix 1. - Updated alcohol-attributable fractions

[Supporting documents (page)](https://fingertips.phe.org.uk/profile/local-alcohol-profiles/supporting-information/supporting-documents)

### Previous

[Updating England-Specific Alcohol-Attributable Fractions (.pdf)](http://www.cph.org.uk/wp-content/uploads/2014/03/24892-ALCOHOL-FRACTIONS-REPORT-A4-singles-24.3.14.pdf) ([local link](`r system.file("aa_documentation", "24892-ALCOHOL-FRACTIONS-REPORT-A4-singles-24.3.14.pdf", package = "aafractions.ncc")`))

* Appendix 2. - Updated alcohol-attributable fractions
    
[Alcohol attributable fractions for England (.pdf)](http://nwph.net/nwpho/publications/alcoholattributablefractions.pdf) ([local link](`r system.file("aa_documentation", "alcoholattributablefractions.pdf", package = "aafractions.ncc")`))

* Table 17. - Alcohol-attributable fractions by age and sex
* Table 20. - AAFs used by the Strategy Unit/Cabinet Office (2003)
* Table 21. - AAFs used to calculate the national indicator for alcohol, NI 39 (provisional figures produced in November 2007)

[HES Alcohol Fractions (.xlsx)](https://digital.nhs.uk/binaries/content/assets/legacy/excel/5/9/hes_apc_data_dictionary_-_supplementary_table.xlsx) ([local link](`r system.file("aa_documentation", "hes_apc_data_dictionary_-_supplementary_table.xlsx", package = "aafractions.ncc")`))

* Worksheet `\Alcohol Fractions/`.


## Smoking attributable fractions

There are two parts: smoking relative risks and smoking prevalence

### Data

[NHS Digital Statistics on Smoking 2018 tables (.xlsx) ](https://files.digital.nhs.uk/5A/FC7B13/stat-smok-eng-2018-tab.xlsx) ([local link](`r system.file("aa_documentation", "stat-smok-eng-2018-tab.xlsx", package = "aafractions.ncc")`))

* Tables in worksheet `\B2/`.

[PHE Local Tobacco Control Profiles  (website)](https://fingertips.phe.org.uk/profile/tobacco-control/data#page/9/gid/1938132886/pat/6/par/E12000004/ati/101/are/E07000170) ([local link](`r system.file("data-raw", "PHE_LTCP_SP_20190304_indicators-DistrictUA.data.csv", package = "aafractions.ncc")`))

* Datasets downloadable from [Downloads] section.

### PHE and NHS Digital methodology

#### Relative risk to Attributable Fraction conversion

[NHS Digital Appendix (.pdf)](https://files.digital.nhs.uk/7A/870C29/stat-smok-eng-2018-app.pdf) ([local link](`r system.file("aa_documentation", "stat-smok-eng-2018-app.pdf", package = "aafractions.ncc")`))

* Appendix section 1.2

#### PHE methodology notes

[PHE Morbidity (web)](https://fingertips.phe.org.uk/profile/tobacco-control/data#page/6/gid/1938132888/pat/6/par/E12000004/ati/102/are/E10000024/iid/1207/age/202/sex/4) ([local link](`r system.file("aa_documentation", "SAF_Admissions_20190304_Public_Health_Profiles.pdf", package = "aafractions.ncc")`))

[PHE Mortality (web)](https://fingertips.phe.org.uk/profile/tobacco-control/data#page/6/gid/1938132887/pat/6/par/E12000004/ati/102/are/E10000024/iid/113/age/202/sex/4) ([local link](`r system.file("aa_documentation", "SAF_Mortality_20190304_Public_Health_Profiles.pdf", package = "aafractions.ncc")`))



## References

PHE (2017), "Local Alcohol Profiles for England 2017 user guide", URL: https://fingertips.phe.org.uk/documents/LAPE_2017_User_Guide_071117.pdf (last accessed 2019 March)

NHS Digital, "Statistics on Smoking", URL: https://digital.nhs.uk/data-and-information/publications/statistical/statistics-on-smoking/ (last accessed 2019 March)

NHS Digital, "Statistics on Smoking 2018", https://digital.nhs.uk/data-and-information/publications/statistical/statistics-on-smoking/statistics-on-smoking-england-2018 (last accessed March 2019)

NHS Digital HES DD (2015), "HES Alcohol Fractions"
, URL: https://digital.nhs.uk/binaries/content/assets/legacy/excel/5/9/hes_apc_data_dictionary_-_supplementary_table.xlsx (last accessed 2019 March)

L. Jones, M. Bellis (2014), "Updating England-Specific Alcohol-Attributable Fractions"
, URL: http://www.cph.org.uk/wp-content/uploads/2014/03/24892-ALCOHOL-FRACTIONS-REPORT-A4-singles-24.3.14.pdf (last accessed 2019 March)

L. Jones, M. Bellis et. al. (2008), "Alcohol attributable fractions for England"
, URL: http://nwph.net/nwpho/publications/alcoholattributablefractions.pdf (last accessed 2019 March)

